Fixing Vulnerabilities Hinders Maintainability

### 2020-01-31 paper/icpc20.tex
new version

Security is a requirement of utmost importance to produce software with 
high quality. However, there is still a considerable amount of 
vulnerabilities being discovered and fixed, almost weekly. We suspect 
that when developers address these vulnerabilities on their codebases, 
they actually affect the maintainability of software, potentially 
introducing other vulnerabilities. This paper evaluates the impact of 
patches to improve security on the maintainability of open-source software. 
Maintainability is measured based on the Better Code Hub’s model of $10$ 
guidelines using a dataset including $1037$ security commits. Results show 
that fixing software vulnerabilities hinders the maintainability of 
open-source software. This is particularly the case for patches that address Path Traversal (CWE-22),
Information Exposure (CWE-200) and Resource Management Errors (CWE-399) . We conclude that changes to code bases 
to fix vulnerabilities need to be performed with extra care.

@@

vulnerabilities in software maintenance accross open-source software.

@@

Maintainability is being calculated as the mean of the guidelines computed
by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on each 
guideline individually. For projects with large 
codebases, the results retrieved for \emph{Keep Your Codebase Small} were not 
viable due to limitations of BCH. Thus, we decided to not consider this guideline 
in our analysis. For each type of
guideline, three types of impact of patches in the software maintainability
are presented: \emph{negative}, maintainability decreases (\emph{red});
\emph{none}, maintainability remains the same (\emph{yellow}); and,
\emph{positive}, maintainability increases (\emph{green}). Next to each
type of change, it is shown the mean ($\overline{x}$) and median (M)
of the maintainability difference and the p-value resulting from the
Paired Wilcoxon signed-rank test.
Regarding the impact of security patches in each guideline, we observe
that overall our patches have a significant positive impact in $5$ guidelines: 
\emph{Write Clean Code} (p-value = $4.9$x$10^{-54}$), \emph{Keep Architecture 
Components Balanced} (p-value = $5.26$x$10^{-71}$), \emph{Couple Architecture 
Components Loosely} (p-value = $2.43$x$10^{-07}$), \emph{Keep Unit Interfaces 
Small} (p-value = $2.25$x$10^{-13}$), \emph{ Write Code Once} (p-value = 
$7.9$x$10^{-35}$). However, each of the previous guidelines also suffer a 
negative impact from $10\%$ to $35\%$. \emph{Separate Concerns in Module} 
guideline also follows the same results but without any statistical 
significance since the p-value is higher than $0.05$. According with the 
results, \emph{Write Simple Units of Code} (p-value=$0.015$ and $38.2\%$/$396$ 
of negative cases) and \emph{Write Short Units of Code} (p-value=$0.029$ and 
$38.3\%$/$397$ of negative cases) are the guidelines which suffered more 
impact from security patches.

@@

We have also evaluated the impact on the maintainability per programming language
of open-source software, Figure~\ref{fig:lang_main}. Although
the dataset contains security patches for more languages, we only
present the results for \emph{Python}, \emph{C/C++}, \emph{Ruby}, \emph{PHP}, 
\emph{Objective-C/C++}, and \emph{Java}
because of the Paired Wilcoxon signed-rank test requirement: input sample has to have more 
than $20$ samples. The other languages were
integrated into \emph{Others}. The results of the Paired Wilcoxon signed-rank
test by language indicate that there is statistical evidence to support that the
maintainability of open-source software decreases for security patches for \emph{C/C++} 
(p-value = $2.5$x$10^{-5}$).
% Following the \textit{American Statistical Association}
% recommendations~\cite{doi:10.1080/00031305.2016.1154108}, we consider the results for
% Ruby as statistically significant because its p-value is on the borderline.
If we considered a significance level of $0.1$, we could also conclude that security 
patches in \emph{PHP} (p-value = $0.072$) also have a negative impact
on the maintainability of software.
However, to the rest of the languages presented in Figure~\ref{fig:lang_main}, it is
not possible to make conclusions with statistical significance. But it is important
to notice that overall there is a very significant number of patches that
hinder the maintainability of software (e.g., \emph{Python}, \emph{Ruby}, 
\emph{Objective-C/C++}, \emph{Java}), even if the number of negative cases does not 
surpass the number of positive or none.

@@

The impact of each type of security patches on the maintainability of
open-source software is presented in Figure~\ref{fig:pat}. The results of the
Paired Wilcoxon signed-rank per pattern exhibited statistical
evidence supporting that the maintainability of open-source software decreases for
only one category: \emph{Miscellaneous} (p-value = $0.033$). Although without 
statistical significance, we also observe that the maintainability of open-source software decreases in $3$ different patterns: \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)}, \emph{Information Exposure (CWE-200)} and \emph{Resource Management Errors (CWE-399)}.

Also with no statistical
significance, results demonstrate evidence that maintainability is less
negatively impacted when patching \emph{SQL Injection (CWE-89)}, \emph{Improper Restriction of XML External Entity Reference (CWE-611)}, \emph{Improper Input Validation (CWE-20)}, \emph{Cross-Site Request Forgery (CWE-352)}, \emph{Permissions, Privileges, and Access Controls (CWE-264)}. However,
for all the previous patterns there is a percentage of security patches that hinder maintainability between $30\%$ and $50\%$ which is very significant. \emph{Cross-Site Scripting (CWE-79)} is the pattern that suffers less positive/negative impact from security patches. As the mean shows, the security
patches of this type of vulnerability usually have no impact on the maintainability of open-source software.

@@

@@

\Sof{Skip, it will be updated soon} For regular commits, we observe that the maintainability decreases in \Sof{<todo>} %$41.35\%$ ($251$) 
of the cases after regular refactorings are applied. But in contrast
to security commits, it increases in \Sof{<todo>} %$58.16\%$ ($353$) 
and remains the same in \Sof{<todo>}
%$0.5\%$ ($3$) 
of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
test is \Sof{<todo>}
%$1.54$x$10^{-6}$
. The baseline is still being computed by BCH but we can see already 
that regular changes usually have no impact on the maintainability of
software which seems to contradict research performed in the past. However,
we inspected a considerable amount of cases where this happens and
found that considering the size of the \emph{diff} to generate the random
regular commit may limit the type of commit obtained. There are several
cases which size is small due to the high amount of

>>>>

Security is a requirement of utmost importance to produce software with 
high quality. However, there is still a considerable amount of 
vulnerabilities being discovered and fixed, almost weekly. We suspect 
that when developers address these vulnerabilities on their codebases, 
they actually affect the maintainability of software, potentially 
introducing other vulnerabilities. This paper evaluates the impact of 
patches to improve security on the maintainability of open-source software. 
Maintainability is measured based on the Better Code Hub’s model of $10$ 
guidelines using a dataset including $1037$ security commits. Results show 
that fixing software vulnerabilities hinders the maintainability of 
open-source software. This is particularly the case for patches that address Path Traversal (CWE-22),
Information Exposure (CWE-200) and Resource Management Errors (CWE-399) . We conclude that changes to code bases 
to fix vulnerabilities need to be performed with extra care and that security-related code evolution should be subject in computer curricula.

@@

vulnerabilities in software maintenance across open-source software.

@@

Maintainability is being calculated as the mean of the guidelines computed
by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on each 
guideline individually. For projects with large 
codebases, the results retrieved for \emph{Keep Your Codebase Small} were not 
viable due to limitations of BCH. For this guideline, the tool sets a limit of 20 Person-years. However, a large amount of projects of our database surpass this limit. Thus, we decided to not consider this guideline 
in our analysis. For each type of
guideline, three types of impact of patches in the software maintainability
are presented: \emph{negative}, maintainability decreases (\emph{red});
\emph{none}, maintainability remains the same (\emph{yellow}); and,
\emph{positive}, maintainability increases (\emph{green}). Next to each
type of guideline, it is shown the mean ($\overline{x}$) and median (M)
of the maintainability difference and the p-value resulting from the
Paired Wilcoxon signed-rank test.
Regarding the impact of security patches in each guideline, we observe
that overall patches have a significant positive impact in $5$ guidelines: 
\emph{Write Clean Code} (p-value = $4.9$x$10^{-54}$), \emph{Keep Arschitecture 
Components Balanced} (p-value = $5.26$x$10^{-71}$), \emph{Couple Architecture 
Components Loosely} (p-value = $2.43$x$10^{-07}$), \emph{Keep Unit Interfaces 
Small} (p-value = $2.25$x$10^{-13}$), \emph{ Write Code Once} (p-value = 
$7.9$x$10^{-35}$). However, for each of these guidelines we also observe a substantial number of cases where the maintainability suffers a negative impact $-$ from $10\%$ to $35\%$ $-$ which highlights the need of better practices/techniques/tools to produce software. \emph{Separate Concerns in Module} 
guideline also follows the same results but without any statistical 
significance since the p-value is higher than $0.05$. According with the 
results, \emph{Write Simple Units of Code} (p-value=$0.015$ and $38.2\%$/$396$ 
of negative cases) and \emph{Write Short Units of Code} (p-value=$0.029$ and 
$38.3\%$/$397$ of negative cases) are the guidelines which suffered more 
impact from security patches.

@@

We have also evaluated the impact on the maintainability per programming language
of open-source software, Figure~\ref{fig:lang_main}. Although
the dataset contains security patches for more languages, we only
present the results for \emph{Python}, \emph{C/C++}, \emph{Ruby}, \emph{PHP}, 
\emph{Objective-C/C++}, and \emph{Java}
because of the Paired Wilcoxon signed-rank test requirement: input sample has to have more 
than $20$ samples. The other languages were
integrated into \emph{Others}. The results of the Paired Wilcoxon signed-rank
test by language indicate that there is statistical evidence to support that the
maintainability of open-source software decreases for security patches for \emph{C/C++} 
(p-value = $2.5$x$10^{-5}$). 
% Following the \textit{American Statistical Association}
% recommendations~\cite{doi:10.1080/00031305.2016.1154108}, we consider the results for
% Ruby as statistically significant because its p-value is on the borderline.
We could also conclude that security 
patches in \emph{PHP} (p-value = $0.072$) also have a negative impact
on the maintainability of software, if we considered a significance level of $0.1$. 
Another programming language which impacts negatively the maintainability of software
is \emph{Ruby} (30 in 71 vulnerabilities, $42.3\%$).

Overall, it is important to notice that there is a very significant number of patches that
hinder the maintainability of software $-$ even if the number of negative cases does not 
surpass the number of positive or none $-$ for  $3$ different languages: \emph{Python},
\emph{Objective-C/C++} and \emph{Java}. For those languages, the negative impact is between $35\%$ and $50\%$ which highlights the need for better and secure programming languages. 
Results should be read with care for the languages where statistical significance was not found. However, 
data reports very interesting hints about the impact of programming languages in 
security patches.

@@

The impact of each type of security patches on the maintainability of
open-source software is presented in Figure~\ref{fig:pat}. Results demonstrate evidence that the maintainability of open-source software decreases in $3$ different patterns: \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)}, \emph{Information Exposure (CWE-200)} and \emph{Resource Management Errors (CWE-399)}.

We also observe that maintainability is less
negatively impacted when patching \emph{SQL Injection (CWE-89)}, \emph{Improper Restriction of XML External Entity Reference (CWE-611)}, \emph{Improper Input Validation (CWE-20)}, \emph{Cross-Site Request Forgery (CWE-352)}, \emph{Permissions, Privileges, and Access Controls (CWE-264)}. However,
for all the previous patterns there is a percentage of security patches that hinder maintainability between $30\%$ and $50\%$ which is very significant. \emph{Cross-Site Scripting (CWE-79)} is the pattern that suffers less positive/negative impact from security patches, i.e., security
patches of this type of vulnerability typically have no impact on the maintainability of open-source software.

@@

The results of the
Paired Wilcoxon signed-rank per pattern exhibited statistical
evidence supporting that the maintainability of open-source software decreases for
only one category: \emph{Miscellaneous} (p-value = $0.033$). However, no statistical
significance was found to the other patterns. Thus, results should be read with care.

@@

For regular commits, we observe that the maintainability decreases in $21.04\%$ ($133$) 
of the cases after regular changes are applied. But in contrast
to security commits, it increases in $26.90\%$ ($170$) 
and remains the same in $52.06\%$ ($329$) 
of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
test is $5.74x10^{-78}$. The baseline is still being computed by BCH but it is possible to observe 
that regular changes typically have little impact on the maintainability which contradicts research performed in the past. However,
we manually inspected a considerable amount of cases where this happens and
found that identifying regular commits as similar in size as the security-related commit is limiting the type of regular commits (e.g, input refactorings, renaming variables or functions)\footnote{Example of a random regular commit generated with our methodology (Accessed on \today{}):
\url{https://github.com/AFNetworking/AFNetworking/compare/9cde4e45845486bdb46ad322448f93df3b8479bf...3486a008a1fd1348d0f2fe748a5e23933d94647a}}.
